Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-08-17 doc/invt.tex
add getav.py

As shown in Table \ref{tab:lj_error},
all updating schemes
with either the inverse-time or optimal schedule
were able to reduced
the error significantly,
and the errors roughly matched the predicted values
calculated using the pre-computed
$\Gamma_k$'s and $\epsilon_k$'s.
%
The Gaussian updating scheme produced less error than
the single-bin scheme in the $\varrho = 0.1$ case,
but not in the $\varrho = 0.8$ case.
%
The effectiveness of the Gaussian updating scheme
in the low-density case
was probably because the sampling along
the collective variable was free from major barriers
and thus the histogram fluctuation largely
came from local noise.
%
%The optimal schedule also appeared to be more effective
%in reducing the error than the inverse-time schedule.
%%
In this case, we could further reduced the error
by using the bandpass updating scheme
(with the mode cutoff $K = 20$).
%
On the other hand, in the high-density case,
the error was dominated by the $k=1$ eigenmode
corresponding to the longest range fluctuation,
as $\Gamma_1 \approx \Gamma$,
and the reduction of local noise
had little effect on the total error.

@@

@@

To illustrate the mode dependence of the error,
we show in Fig. \ref{fig:lj_xerr}
the error components, $\langle \tilde v_k^2 \rangle$,
against the eigenmode index, $k$,
for the $\varrho = 0.1$ case
with width $\sigma = 0.2$.
%
For modes $k=10$ to $25$,
the final errors from the Gaussian updating scheme
were greater than those from the single-bin updating scheme,
although the initial values were less.
%
For the Gaussian updating scheme,
the inverse-time schedule tended to
reduce the updating magnitude too quickly,
leaving a hump on the error profile
for these middle-range eigenmodes.
%
The optimal schedule improved over the inverse-time schedule
and produced a flatter profile for the final error.
%
As a result, the optimal schedule usually produced
less error than the inverse-time schedule,
as shown in Table \ref{tab:lj_error}.
%
Although the gain of using the optimal schedule
was modest in the reported cases,
it would increase with the simulation length.
%
For example, for the $\varrho = 0.1$ case
with width $\sigma = 0.2$,
if we extend the simulation to $10^8$ steps,
the error for the optimal schedule was
$2.9\times10^{-6}$,
which was much smaller than that
for the inverse-time schedule
$1.8\times10^{-5}$.
\note{The predicted errors were
$3.7\times10^{-6}$ and $1.8\times10^{-5}$, respectively.}

@@

In Fig. \ref{fig:lj_alpha},
we show the optimal schedules in a few cases.
%
%Generally, the optimal schedule
%for the $\varrho = 0.8$ case
%was closer to the inverse-time schedule
%than that for the $\varrho = 0.1$ case.
%%
%This was probably because in the former case
%the first mode dominates the total error,
%$\Gamma_1 \approx \Gamma$,
%and the schedule was roughly optimized for this mode,
%resulting in
%$\alpha(t) \approx 1/[\lambda_1 (t + t_0)] \approx 1/(t+t_0)$
%as $\lambda_1 \approx 1$
%[cf. Eq. \eqref{eq:lambda_Gaussian}].
%
Since we have chosen a long equilibration period
for the preliminary run under the constant updating magnitude, $a_0$,
the initial magnitude was around
the ideal value of $a_0/2 = 5\times10^{-5}$
given by Eq. \eqref{eq:half_alpha0}.
%
But if the Gaussian were too wide,
e.g. if $\sigma = 0.6$,
the eigenvalues $\lambda_k$ of some modes
became too small to damp out the initial bias,
resulting in a much larger initial updating magnitude, $\alpha(0)$
[cf. Eq. \eqref{eq:opt_alpha0}].
%
In this case, the predicted schedule
would be unusable because the resulting deviation in the bias potential
was no longer small,
and one may have to either decrease the width
or to elongate the preliminary run.

>>>>

As shown in Table \ref{tab:lj_error},
both the optimal and inverse-time schedules
were able to reduced the error significantly,
suggesting the potential benefit of using a decreasing updating magnitude
even with a sometimes sub-optimal schedule.
%
It is also reassuring to see that
the errors roughly matched the predicted values
calculated using the pre-computed
$\Gamma_k$'s and $\epsilon_k$'s.
%
Since the two schedules are identical
in the single-bin and bandpass updating schemes,
we will focus on the Gaussian updating scheme below
in discussing the optimal schedule.

@@

As shown in Fig. \ref{fig:lj_alpha},
the optimal schedules for the Gaussian updating schemes
differed from inverse-time one,
and the former tended to be larger.
%
The optimal schedule was more effective
than the sub-optimal inverse-time schedule
in reducing the error,
as shown in Table \eqref{tab:lj_error}.
%
Although the advantage of the former
was modest in the reported cases,
it would increase with the simulation length.
%
For example, for the case of
$\varrho = 0.1$ and Gaussian width $\sigma = 0.2$,
if we extend the simulation to $10^8$ steps,
the error for the optimal schedule was
$2.9\times10^{-6}$,
which was much smaller than that
for the inverse-time schedule
$1.8\times10^{-5}$.
  \note{The predicted errors were
  $3.7\times10^{-6}$ and $1.8\times10^{-5}$, respectively.}

To illustrate the mode dependence of the error,
we show in Fig. \ref{fig:lj_xerr}
the error components, $\langle \tilde v_k^2 \rangle$,
against the eigenmode index, $k$,
for the $\varrho = 0.1$ case
with width $\sigma = 0.2$.
%
For modes $k=10$ to $25$,
the final errors from the Gaussian updating scheme
were greater than those from the single-bin updating scheme,
although the initial values were less.
%
For the Gaussian updating scheme,
the inverse-time schedule tended to
reduce the updating magnitude too quickly,
leaving a hump on the error profile
for these middle-range eigenmodes.
%
The optimal schedule improved over the inverse-time schedule
and produced a flatter profile for the final error.
%
As a result, the optimal schedule usually produced
less error than the inverse-time schedule,
as shown in Table \ref{tab:lj_error}.
%

@@

@@

The Gaussian updating scheme produced less error than
the single-bin scheme in the $\varrho = 0.1$ case,
but not in the $\varrho = 0.8$ case.
%
The effectiveness of the Gaussian updating scheme
in the low-density case
was probably because the sampling along
the collective variable was free from major barriers
and thus the histogram fluctuation largely
came from local noise.
%
%The optimal schedule also appeared to be more effective
%in reducing the error than the inverse-time schedule.
%%
In this case, we could further reduced the error
by using the bandpass updating scheme
(with the mode cutoff $K = 20$).
%
On the other hand, in the high-density case,
the error was dominated by the $k=1$ eigenmode
corresponding to the longest range fluctuation,
as $\Gamma_1 \approx \Gamma$,
and the reduction of local noise
had little effect on the total error.

But if the Gaussian were too wide,
e.g. if $\sigma = 0.6$,
the eigenvalues $\lambda_k$ of some modes
became too small to damp out the initial bias,
resulting in a much larger initial updating magnitude, $\alpha(0)$
[cf. Eq. \eqref{eq:opt_alpha0}].
%
In this case, the predicted schedule
would be unusable because the resulting deviation in the bias potential
was no longer small,
and one may have to either decrease the width
or to elongate the preliminary run.

The initial magnitude was, however, around
the ideal value of $a_0/2 = 5\times10^{-5}$
given by Eq. \eqref{eq:half_alpha0}.
%
Since we have chosen a long equilibration period
for the preliminary run under the constant updating magnitude, $a_0$,
